dyngen: simulating developing single cells

### 2019-09-18 content.tex
update

@@

One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be simulated. Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies. Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments. % could use figure; one with GRNs of different topologies, another with rewiring events
Multiple simulations with some slight perturbations can be used to replicate batch effects.

@@

@@

%% TODO: show that dyngen output resembles real data using e.g. countsimQC?

%% TODO: add result figures pertaining use cases

@@

Currently, dyngen focuses on simulating cells as standalone entities.
Future developments include extending the framework to simulate multiple cells in a virtual environment. Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration. 
%¡also adding more types of molecules, e.g. protein complex, small rnas, PPI
\section{Methods}
NOTE: The method section is really rough at this stage.

@@

TODO: mention strength and cooperativity.

@@

\subsection{Generate gene regulatory network}
% TODO: write section in words instead of bullet point list

@@

\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
% TODO: write section in words instead of bullet point list

Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.

@@

% TODO: explain derivation of transcription formula? 
% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well. 
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independent and result in multiplicative effect.

@@

% todo ask Wouter, why not simply the following?
% ba_G = ifelse(R^+_G is not empty, 1, 0)

@@

% → basically an ODE
When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state. It is assumed the root state has no modules active and has no expression of any molecules. To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition. To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.

@@

Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.

>>>>

\comment{GRN is defined nowhere}

@@

One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be simulated. Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies. Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments. 
Multiple simulations with some slight perturbations can be used to replicate batch effects.

\comment{could use figure; one with GRNs of different topologies, another with rewiring events}

@@

\comment{explain what is meant by 'sampling' and 'profiling technique' a little better}

@@

\comment{show that dyngen output resembles real data using e.g. countsimQC?}

\comment{add result figures pertaining use cases}

@@

Currently, dyngen focuses on simulating cells as standalone entities.
Future developments include extending the framework to simulate multiple cells in a virtual environment. Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.

\comment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
\section{Methods}
\comment{The method section is REALLY rough at this stage.}

@@

\comment{mention strength and cooperativity.}

@@

\subsection{Generate gene regulatory network}
\comment{Welcome to the beginning of the end}

@@

\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
Each reaction consists of its propensity -- a formula to calculate
the probability of the reaction occurring during an infinitesimal
time interval -- and the effect -- how it will
 affect the current state if triggered.

@@

\comment{TODO: add step-by-step  derivation of transcription formula}
% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well. 
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independent and result in multiplicative effect.

@@

@@

\comment{→ basically an ODE}
When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state. It is assumed the root state has no modules active and has no expression of any molecules. To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition. To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.

@@

Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.
